Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-04 submission.tex
Added results from parsing with held-out

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{acl2016}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{booktabs}
\usepackage{relsize}

%\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

@@

\section{Introduction}
\label{sec:intro}
Part-of-speech (PoS) tagging is an important preprocessing step for many NLP tasks,
such as dependency parsing \cite{Niv:Hal:Kub:07,Haj:Cia:Joh:09}, named
entity recognition \cite{Tjo:DeM:03} and sentiment analysis
\cite{Wil:Wie:Hof:09}. Whereas much effort has gone into the
development of PoS-taggers -- to the effect that this task is often
considered more or less a solved task -- considerably less effort has
been devoted to the empirical evaluation of the PoS tag sets
themselves. Error analysis of PoS-taggers indicate that, whereas
tagging improvement through means of learning algorithm or feature
engineering seems to have reached something of a plateau, linguistic
and empirical assessment of the distinctions made in the PoS tag set may be a venue
worth investigating further \cite{Man:11}.  Clearly, the utility of a
PoS tag set is tightly coupled with the the downstream task for which
it is performed. Even so, PoS tag sets are usually employed in a ``one
size fits all'' fashion, regardless of the requirements posed by the
task which makes use of this information.

@@

In this paper, we optimize a PoS tag set for the task of
data-driven dependency parsing of Norwegian. We report on a set of experiments where various morphological
distinctions are introduced to the PoS annotations and evaluated both intrinsically, i.e. in
terms of PoS-tagging accuracy, and extrinsically, in terms of parsing
accuracy.  Our results show that the introduction of morphological
distinctions not present in the original tag set, whilst compromising
tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a
simple tagging task.

@@

\section{Previous/Related work}
\label{sec:prev}
There has been little previous work dedicated specifically to
extrinsic evaluation of the effects of PoS tag sets on downstream
applications.  There has, however, been some work that evaluates the
effects of PoS tag set granularity on PoS tagging. This includes
investigation of the effects of PoS tag sets on tagging of Swedish
\cite{Meg:01,Meg:02} and English \cite{Mac:05}.

@@

\newcite{Meg:01} and \newcite{Meg:02} trained and evaluated a range of PoS
taggers on the Stockholm-Umeå Corpus (SUC) \cite{Gus:Har:06}, annotated with a
tag set based on a Swedish version of PAROLE tags totaling 139 tags.
Furthermore, they investigated the effects of tag set size on tagging by
mapping the original tag set into smaller subsets designed for parsing. They
argue that a tag set with complete morphological tags may not be necessary for
all NLP applications, for instance syntactic parsing. They found that the
smallest tag set comprising 26 tags yields the lowest tagger error rate.
However, for some of the taggers, augmenting the tag set with more
linguistically informative tags may actually lead to a drop in error rate. They
argue that this shows that the size of the tag set as well as the type of
information in the tags are crucial factors for tagger performance.
However/unfortunately, they do not report results of parsing with the various
PoS tag sets.

@@

@@

\paragraph{Tags \& Features}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we want to evaluate the parser on automatically
assigned PoS tags. For the training of the parser, however, we have two
options: using either gold standard or automatically assigned tags. In order to
settle on a configuration, we conducted experiments with gold standard and
automatically assigned tags to see how they differ with respect to performance.
The results of our experiments reveal that the combination of training and
testing on automatic tags is superior to training on gold standard tags and
testing on automatic tags.  Consequently, the parser was both trained and
tested on automatically assigned tags in our experiments.

@@

In the following, we report on a set of experiments which evaluate
increasingly fine-grained PoS tag sets. The PoS tag sets are motivated
by morphosyntactic properties of Norwegian and are evaluated both in
terms of tagger and parser accuracy.

\section{Tag Set Optimization/Experiments}
\label{sec:optimization}
By introducing more fine-grained linguistically motivated distinctions in a tag
set, we increase the linguistic information represented in the tags, which may
assist the parser in recognizing and generalizing syntactic patterns. However,
the addition of more linguistic information to the tags and thus a more
fine-grained tag set will most likely lead to a drop in tagger accuracy/tagging
quality due to the increase in complexity. The best tagging does not
necessarily lead to the best parse, and it is therefore interesting to
investigate how the tag set modifications may affect the interplay between
tagging and parsing. As we want to investigate how we can improve the
linguistic quality of a tag set, we will not introduce tag set modifications
which are not linguistically motivated. We will only consider distinctions we
deem linguistically sensible, even if we expect them to impair the
computational tractability; investigating the interplay between these two
considerations is ultimately our goal.

\subsection{Initial/Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). The result of this was a total of 368 tags, which is clearly very
fine-grained. The two initial tag sets, i.e., the original tag set comprising
19 tags and the full tag set comprising 368 tags, thus represent two extremes
in terms of granularity.  To evaluate these tag sets and investigate how the tag
set granularity would affect the performance of tagging and parsing, we trained
and evaluated TnT and Mate on the training and development data, respectively,
on the two tag sets. In Table \ref{inittagseteval}, we report the results of
these experiments. We see that the tagger accuracy drastically drops when going
from the original to the full tag set. TnT reports an accuracy of 97.47\% on
the original tag set, which is reduced to 93.48\% for the full tag set. These
results confirm our hypothesis that the very high linguistic quality in the
full, fine-grained tag set comes at the expense of drops in tagger performance.
However, the additional linguistic information provided by the full tag set
improves the parser performance. With the original tag set, Mate reports an LAS
of 87.01\% and a UAS of 90.19\%, which increases to 87.15\% and 90.39\%,
respectively, when using the full tag set. As we are looking for linguistically
informed distinctions that improve the syntactic parsing, these results are
promising and serve to indicate that additional morphological information
assists the syntactic parsers, motivating the optimization of the existing PoS
tag set.

@@

\subsection{Experiments}
"Nearly" all tag set modifications leads to a drop in tagger accuracy, as
expected. Few exceptions...

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Feature} & \textbf{MFT} & \textbf{Accuracy} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & \textbf{94.14\%} & 97.47\% & 87.01\% & 90.19\% \\
        Definiteness & 94.13\% & 97.49\% & \textbf{87.30\%} &
        \textbf{90.42\%} \\
        Gender & 93.86\% & 97.28\% & 87.09\% & 90.31\% \\
        Number & 94.06\% & 97.49\% & 87.04\% & 90.18\% \\
        Type & 94.14\% & \textbf{97.61\%} & 87.00\% & 90.11\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for determiners.}
    \label{detresults}
\end{table}

@@

@@

@@

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Tag set} & \textbf{MFT} & \textbf{Accuracy} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        Optimized & 89.20\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets.}
    \label{finalresults}
\end{table}

@@

\section{Summary/Conclusion and Future Work}
\label{sec:summary}

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{acl2016}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{booktabs}
\usepackage{relsize}
\usepackage{multirow}
%\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

@@

\section{Introduction}
\label{sec:intro}
Part-of-speech (PoS) tagging is an important preprocessing step for many NLP
tasks, such as dependency parsing \cite{Niv:Hal:Kub:07,Haj:Cia:Joh:09}, named
entity recognition \cite{Tjo:DeM:03} and sentiment analysis
\cite{Wil:Wie:Hof:09}. Whereas much effort has gone into the development of
PoS-taggers -- to the effect that this task is often considered more or less a
solved task -- considerably less effort has been devoted to the empirical
evaluation of the PoS tag sets themselves. Error analysis of PoS taggers
indicate that, whereas tagging improvement through means of learning algorithm
or feature engineering seems to have reached something of a plateau, linguistic
and empirical assessment of the distinctions made in the PoS tag sets may be a
venue worth investigating further \cite{Man:11}. Clearly, the utility of a PoS
tag set is tightly coupled with the downstream task for which it is performed.
Even so, PoS tag sets are usually employed in a ``one size fits all'' fashion,
regardless of the requirements posed by the task which makes use of this
information.

@@

In this paper, we optimize a PoS tag set for the task of data-driven dependency
parsing of Norwegian. We report on a set of experiments where various
morphological distinctions are introduced to the PoS annotations and evaluated
both intrinsically, i.e., in terms of PoS tagging accuracy, and extrinsically,
in terms of parsing accuracy. Our results show that the introduction of
morphological distinctions not present in the original tag set, whilst
compromising tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a simple
tagging task.

@@

\section{Previous Work}
\label{sec:prev}
There has been little previous work dedicated specifically to
extrinsic evaluation of the effects of PoS tag sets on downstream
applications.  There has, however, been some work that evaluates the
effects of PoS tag set granularity on PoS tagging. This includes
investigation of the effects of PoS tag sets on tagging of Swedish
\cite{Meg:01,Meg:02} and English \cite{Mac:05}.

@@

\newcite{Meg:02} trained and evaluated a range of PoS taggers on the
Stockholm-Umeå Corpus (SUC) \cite{Gus:Har:06}, annotated with a tag set based
on a Swedish version of PAROLE tags totaling 139 tags.  Furthermore, they
investigated the effects of tag set size on tagging by mapping the original tag
set into smaller subsets designed for parsing. They argue that a tag set with
complete morphological tags may not be necessary for all NLP applications, for
instance syntactic parsing. They found that the smallest tag set comprising 26
tags yields the lowest tagger error rate.  However, for some of the taggers,
augmenting the tag set with more linguistically informative tags may actually
lead to a drop in error rate. They argue that this shows that the size of the
tag set as well as the type of information in the tags are crucial factors for
tagger performance.  However/unfortunately, they do not report results of
parsing with the various PoS tag sets.

@@

Our approach is somewhat semi-automatic in that we use a "mix" of both manual
annotation in our experiments. Our initial tags are gold standard, and using
linguistic insight coupled with computational considerations, we introduce new
features or feature combination to the tag set. These tags are run through a
pipeline with automatically assigned tags in both training and testing, where
the most promising modifications are used "further". Hence, the tag set
modifications are not deterministic, nor do we try all possible combination; we
only experiment with linguistically motivated distinctions.

@@

\paragraph{Tags \& Features}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we want to evaluate the parser on automatically
assigned PoS tags. For the training of the parser, however, we have two
options: using either gold standard or automatically assigned tags. In order to
settle on a configuration, we conducted experiments with gold standard and
automatically assigned tags to see how they differ with respect to performance.
The results of our experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on automatic tags is clearly superior
to training on gold standard tags and testing on automatic tags. Consequently,
the parser was both trained and tested on automatically assigned tags in our
experiments.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Training} & \textbf{Testing} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Gold & Gold & 90.15\% & 92.51\% \\
        Gold & Auto & 85.68\% & 88.98\% \\
        Auto & Auto & 87.01\% & 90.19\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of parsing with Mate using the various tag configurations,
        either using gold standard tags or automatically assigned from TnT.
        \emph{Gold} denotes gold standard tags, \emph{Auto} denotes
        automatically assigned tags from TnT.}
    \label{tagconfigeval}
\end{table}

@@

In the following, we report on a set of experiments which evaluate increasingly
fine-grained PoS tag sets. The PoS tag sets are motivated by morphosyntactic
properties of Norwegian and are evaluated in terms of both tagger and parser
accuracy.

\section{Tag Set Optimization}
\label{sec:optimization}
By introducing more fine-grained linguistically motivated distinctions in a tag
set, we increase the linguistic information represented in the tags, which may
assist the parser in recognizing and generalizing syntactic patterns. However,
the addition of more linguistic information to the tags and thus a more
fine-grained tag set will most likely lead to a drop in tagger accuracy/tagging
quality due to the increase in complexity. The best tagging does not
necessarily lead to the best parse, and it is therefore interesting to
investigate how the tag set modifications may affect the interplay between
tagging and parsing.

As we want to investigate how we can improve the
linguistic quality of a tag set, we will not introduce tag set modifications
which are not linguistically motivated. We will only consider distinctions we
deem linguistically sensible, even if we expect them to impair the
computational tractability; investigating the interplay between these two
considerations is ultimately our goal.

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). This resulted in a total of 368 tags, which is clearly very fine-grained.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity.  To evaluate these tag sets and investigate how the tag set
granularity would affect the performance of tagging and parsing, we trained and
evaluated TnT and Mate on the training and development data, respectively, on
the two tag sets. In Table \ref{inittagseteval}, we report the results of these
experiments. We see that the tagger accuracy drastically drops when going from
the original to the full tag set. TnT reports an accuracy of 97.47\% on the
original tag set, which is reduced to 93.48\% for the full tag set. These
results confirm our hypothesis that the very high linguistic quality in the
full, fine-grained tag set comes at the expense of drops in tagger performance.
However, the additional linguistic information provided by the full tag set
improves the parser performance. With the original tag set, Mate reports an LAS
of 87.01\% and a UAS of 90.19\%, which increases to 87.15\% and 90.39\%,
respectively, when using the full tag set. As we are looking for linguistically
informed distinctions that improve the syntactic parsing, these results are
promising and serve to indicate that additional morphological information
assists the syntactic parsers, motivating the optimization of the existing PoS
tag set.

@@

\subsection{Experiments}
"Nearly" all tag set modifications leads to a drop in tagger accuracy, as
expected. Few exceptions... We observed increases in parser accuracy scores
with linguistically motivated tag set modifications for all PoS tags we
experimented with, the most improved being noun with an increase in LAS from
the original tag set by 1.80 percentage points.

@@

For a number of reasons, we do not report results from combinations of features
for determiners, as the features could not be combined in any meaningful way.
For instance, determiners in plural never have marked definiteness, hence this
combination would only apply to determiners in singular and roughly correspond
to the distinction on definiteness alone. Moreover, determiners in plural are
not distinguished in terms of gender (thus ruling out the combination of number
and gender), neither are definite determiners. Another aspect taken into
consideration is that the most promising distinction, definiteness, applies to
such a small number of tokens that more fine-grained distinctions would be
overly sparse.

The results from the experiments with determiners are shown in Table
\ref{detresults}.  Introducing the type led to an increase in tagger accuracy
by 0.14 percentage points to 97.61\%, while marginally impacting the parsing,
with LAS of 87.00\%, 0.01 percentage points below the baseline, and UAS of
90.11\%, 0.08 percentage points below the baseline.
%The fact that the baseline
%remains the same is especially worth noting, as increased tag set granularity
%almost inevitably leads to a drop in the MFT baseline.
The increase in tagger
accuracy when introducing the distinction of type is noteworthy, as we expected
the finer granularity to lead to a decrease in accuracy. This serves to
indicate that more fine-grained distinctions for determiners, which is a quite
disparate category in the treebank, may be quite useful for tagging. However,
as it has negative impact on the syntactic parsing, we can conclude that the
type of a determiner does not assist in generalizing syntactic patterns, as
most determiners, regardless of type, appear in the same syntactic
constructions (i.e., before a noun or noun phrase).

Gender, on the other hand, improved the parsing, but complicated the tagging,
as the various genders are often difficult to differentiate, especially so in
the case of masculine and feminine, which share many of the same determiners.
The number of a determiner, i.e., singular or plural, led to a small increase
in tagger accuracy and LAS, while marginally lower UAS, 90.18\%, 0.01
percentage points lower than that of the original tag set. Almost all
determiners have number, and the introduction of this distinction led to small
increases in tagger accuracy and LAS, but marginally lower UAS. The
introduction of definiteness to the determiners led to the best parsing
results, LAS of 87.30\% and UAS of 90.42\%, while also increasing the tagger
accuracy slightly. The increase in LAS and UAS is rather interesting, as there
are only 121 "determiner tokens" with marked definiteness in the development
data. As this accounts for a very small number of tokens, coupled with the
previously noted considerations, we did not consider further fine-grained
modifications with definiteness. This goes to show that tokens with overt
definiteness have noticeable impact on the syntactic parsing, and that
distinguishing on definiteness is very beneficial.

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & 97.47\% & 87.01\% & 90.19\% \\
        %Definiteness & 94.13\% & 97.49\% & \textbf{87.30\%} &
        %\textbf{90.42\%} \\
        %Gender & 93.86\% & 97.28\% & 87.09\% & 90.31\% \\
        %Number & 94.06\% & 97.49\% & 87.04\% & 90.18\% \\
        %Type & 94.14\% & \textbf{97.61\%} & 87.00\% & 90.11\% \\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for determiners.}
    %\label{detresults}
%\end{table}

@@

The results in Table \ref{pronresults} show that number, person and type are
the most informative features for parsing, with LAS of 87.21\%, 87.22\% and
87.19\%, respectively. However, when combining number and person, we observe a
drop by more than 0.2 percentage points, indicating that these features do not
interact in any syntactically distinctive way. The most interesting observation
is that all experiments exceed the baseline tagger accuracy, the most improved
being the most fine-grained distinction, namely type, case and number combined,
obtaining a tagger accuracy of 97.52\%. This shows that the introduction of
more fine-grained distinctions for pronouns is beneficial and aids the PoS
tagger in disambiguating ambiguous words. While case alone yields an LAS of
87.08\%, we found that the combination of type and case, which is the most
successful experiment in terms of parser performance, yields the second highest
tagging accuracy of 97.51\%. The reason for this is that pronouns of different
type and personal pronouns of different case exhibit quite different properties
and appear in different constructions. Pronouns in nominative case (i.e.,
subjects) primarily occur before the main verb, while pronouns in accusative
case (i.e., objects) occur after the main verb, as Norwegian exhibits so-called
V2 word order, requiring that the finite verb of a declarative clause appears
in the second position, hence its name. The combination of type and number
comes in close to the performance of type and case, with LAS of 87.27\% and UAS
identical to that of type and case.

@@

\subsection{Optimized Tag Set}
The most successful tag set modifications and their results are seen in Table
\ref{respectiveresults}.

@@

An overview of the tags comprising the final, optimized tag set can be seen in
Table \ref{optimizedtagset}.

@@

Results from parsing with the optimized tag are shown in Table
\ref{finalresults}, compared to the results obtained with the initial tag sets.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Data Set} & \textbf{Tag Set} & \textbf{MFT} & \textbf{Accuracy}
        & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 89.20\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & 94.22\% & 97.38\% & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 88.08\% & 96.35\% & 88.55\% & 91.41\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets. Trained and tested using
        automatically assigned tags from TnT.}
    \label{finalresults}
\end{table}

@@

\section{Summary and Future Work}
\label{sec:summary}
The improvements in parser performance with our optimized tag set indicate that
a more fine-grained PoS tag set may assist syntactic parsers in recognizing and
generalizing syntactic patterns, while potentially compromising the performance
of PoS taggers.

There are several aspects of this thesis that can be further explored in future
work, including extrinsic evaluation of the effects of PoS tag sets on other
downstream NLP applications besides parsing, such as sentiment analysis and
named entity recognition. These applications often require tagged data, but are
markedly different from syntactic parsing, hence the evaluation would involve
investigating an entirely different aspect of the effects of tag set
granularity.
